---
title: "Data_analysis"
author: "Kristi Fourie"
date: "2024-05-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Analysis

This is a R markdown document where we undergo scRNA seq data analysis of regulatory T cells from tumor and healthy tissue

```{r setup}
# Load necessary libraries
library(EnsDb.Hsapiens.v86)  # For mapping Ensembl IDs to gene symbols
library(AnnotationDbi)       # Provides the `mapIds` function
library(readr)               # For reading CSV files
library(dplyr)               # For data manipulation
library(Seurat)  
library(ggplot2) 
library(pheatmap)  
library(DESeq2)
library(tidyr)
library(tibble)
```

## Load the counts data 
```{r load_data}
# Read tumor and healthy count matrices as tibbles
counts_tumor <- read_csv('./data_tumor_atlas/tumor_raw_counts.csv', col_names = TRUE, show_col_types = FALSE)
counts_health <- read_csv('./data_healthy_atlas/healthy_raw_counts.csv', col_names = TRUE, show_col_types = FALSE)

# Rename the first column to 'Geneid'
colnames(counts_tumor)[1] <- 'Geneid'
colnames(counts_health)[1] <- 'Geneid'

# Convert tibbles to data frames
counts_tumor <- as.data.frame(counts_tumor)
counts_health <- as.data.frame(counts_health)

# Set the 'Geneid' column as the row names
rownames(counts_tumor) <- counts_tumor$Geneid
counts_tumor <- counts_tumor[,-1]

rownames(counts_health) <- counts_health$Geneid
counts_health <- counts_health[,-1]
```
## Mapping Ensembl IDs to Gene Symbols

```{r mapping}
# Map Ensembl IDs to gene symbols for the healthy dataset
ens <- mapIds(EnsDb.Hsapiens.v86,
              keys = rownames(counts_health),
              column = 'SYMBOL',
              keytype = 'GENEID',
              multiVals = "first")

# Filter out NA values and update row names
keep <- !is.na(ens)
ens <- ens[keep]
counts_health <- counts_health[keep, ]
```

```{r unique_symbols}
# Function to ensure unique gene symbols
make_unique <- function(symbols) {
  duplicates <- symbols[duplicated(symbols) | duplicated(symbols, fromLast = TRUE)]
  symbol_counts <- table(duplicates)
  
  symbols_out <- symbols
  for (dup in names(symbol_counts)) {
    dup_indices <- which(symbols == dup)
    suffix <- seq_along(dup_indices)
    symbols_out[dup_indices] <- paste0(dup, "_", suffix)
  }
  return(symbols_out)
}

# Update row names with unique gene symbols
rownames(counts_health) <- make_unique(ens)
```

## Post mapping seurat objects
```{r seurat_object}
# Create Seurat objects from filtered data
tumor_seurat <- CreateSeuratObject(counts = counts_tumor, project = "Tumor", min.cells = 3, min.features = 200)
healthy_seurat <- CreateSeuratObject(counts = counts_health, project = "Healthy", min.cells = 3, min.features = 200)
```

## Differential Gene Expression
```{r deseq2}
# Load required libraries
library(DESeq2)
library(tidyr)
library(tibble)

# Combine tumor and healthy data into one data frame and add metadata
combined_counts <- cbind(tumor_counts, healthy_counts)
combined_metadata <- data.frame(
  sample = colnames(combined_counts),
  condition = c(rep("Tumor", ncol(tumor_counts)), rep("Healthy", ncol(healthy_counts)))
)

# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = combined_counts,
                              colData = combined_metadata,
                              design = ~ condition)

# Run DESeq2 analysis
dds <- DESeq(dds)

# Get results
res <- results(dds)
res <- res[order(res$padj, na.last = NA), ]

# Save results to a CSV file
write.csv(as.data.frame(res), file = "deseq2_results.csv")
```

## Process Data
Data gets processed again to ensure that it is correctly scaled and normalized
```{r process_data}
# Normalize data
tumor_seurat <- NormalizeData(tumor_seurat, normalization.method = "LogNormalize", scale.factor = 10000)
healthy_seurat <- NormalizeData(healthy_seurat, normalization.method = "LogNormalize", scale.factor = 10000)

# Find variable features
tumor_seurat <- FindVariableFeatures(tumor_seurat, selection.method = "vst", nfeatures = 2000)
healthy_seurat <- FindVariableFeatures(healthy_seurat, selection.method = "vst", nfeatures = 2000)

# Scale data
tumor_seurat <- ScaleData(tumor_seurat)
healthy_seurat <- ScaleData(healthy_seurat)
```

## Dimensionality Reduction and Clustering
```{r clustering_pca}
# Perform PCA
tumor_seurat <- RunPCA(tumor_seurat, features = VariableFeatures(object = tumor_seurat))
healthy_seurat <- RunPCA(healthy_seurat, features = VariableFeatures(object = healthy_seurat))

# Run UMAP
tumor_seurat <- RunUMAP(tumor_seurat, dims = 1:30)
healthy_seurat <- RunUMAP(healthy_seurat, dims = 1:30)

# Find clusters using Leiden algorithm
tumor_seurat <- FindNeighbors(tumor_seurat, dims = 1:30)
tumor_seurat <- FindClusters(tumor_seurat, resolution = 0.5)

healthy_seurat <- FindNeighbors(healthy_seurat, dims = 1:30)
healthy_seurat <- FindClusters(healthy_seurat, resolution = 0.5)
```

## Visualizing Clusters
```{r visualise}
# UMAP plots
DimPlot(tumor_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Tumor Clusters")
DimPlot(healthy_seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Healthy Clusters")
```

## Feature mapping
```{r visualise}
# Include additional data to display alongside cell names by passing in a data frame of
# information.  Works well when using FetchData
plot1 <- FeaturePlot(tumor_seurat, features = "FOXP3")
plot2 <- FeaturePlot(tumor_seurat, features = "IL2RA")
plot3 <- FeaturePlot(healthy_seurat, features = "FOXP3")
plot4 <- FeaturePlot(healthy_seurat, features = "IL2RA")
plot1 +  ggmin::theme_powerpoint() + labs(title = "FOXP3 Expression in TME")
plot2 +  ggmin::theme_powerpoint() + labs(title = "CD25 Expression in TME")
plot3 +  ggmin::theme_powerpoint() + labs(title = "FOXP3 Expression in Breast Tissue")
plot4 +  ggmin::theme_powerpoint() + labs(title = "CD25 Expression in Breast Tissue")
#HoverLocator(plot = plot, information = FetchData(tumor_seurat, vars = c("ident", "PC_1", "nFeature_RNA")))
```

## Dot Plots of Gene Groups
```{r dotplots}
stress_genes_features <- c('HSPA1A', 'HSPA1B', 'DNAJB1', 'DNAJA1', 'HSPA8')
inflammatory_genes_features <- c('IFITM1', 'DUSP1', 'CD74', 'RAC2', 'DUSP4', 'CXCL13')
cytoskeleton_genes_features <- c('ACTB', 'PFN1')
metabolic_genes_features <- c('GAPDH', 'ALDOA', 'PKM')
transcription_factor_genes_features <- c('FOS', 'JUN')

gene_group = stress_genes_features ##change here accordingly

# Create the DotPlot and customize the axes labels
DotPlot(tumor_seurat, features = gene_group) + 
  RotatedAxis() + 
  ggmin::theme_powerpoint() + 
  labs(x = "Genes", 
       y = "Clusters")

DotPlot(healthy_seurat, features = gene_group) + 
  RotatedAxis() + 
  ggmin::theme_powerpoint() + 
  labs(x = "Genes", 
       y = "Clusters")
```
## Extracting expression values
```{r expression_vals}
# Extract the expression data for the genes of interest
expression_data_tumor <- FetchData(tumor_seurat, vars = identifiers)
expression_data_healthy <- FetchData(healthy_seurat, vars = identifiers)

# Add cluster information to the data
expression_data_tumor$cluster <- Idents(tumor_seurat)
expression_data_healthy$cluster <- Idents(healthy_seurat)

# Reshape the data using melt from reshape2
melted_data_tumor <- melt(expression_data_tumor, id.vars = "cluster", variable.name = "gene", value.name = "expression")
melted_data_healthy <- melt(expression_data_healthy, id.vars = "cluster", variable.name = "gene", value.name = "expression")

# Calculate the percentage of cells expressing each gene in each cluster
percentage_expressed_tumor <- melted_data_tumor %>%
  group_by(cluster, gene) %>%
  summarise(
    percent_expressed = sum(expression > 0) / n() * 100,
    average_expression = mean(expression)
  )

percentage_expressed_healthy <- melted_data_healthy %>%
  group_by(cluster, gene) %>%
  summarise(
    percent_expressed = sum(expression > 0) / n() * 100,
    average_expression = mean(expression)
  )

# Display the results
print(percentage_expressed_tumor)
print(percentage_expressed_healthy)
```
```{r average_expression}
# Calculate the average expression of each gene across all clusters
average_expression <- percentage_expressed_healthy %>%
  group_by(gene) %>%
  summarize(avg_percent_expressed = mean(percent_expressed, na.rm = TRUE))

# Print the result
print(average_expression)
```

## Find Markers
```{r find_markers}
tumor_seurat.markers <- FindAllMarkers(tumor_seurat, only.pos = TRUE)
healthy_seurat.markers <- FindAllMarkers(healthy_seurat, only.pos = TRUE)
```

## Average Expression per cluster
```{r combine}
# Extract count data and cluster information for tumor samples
tumor_counts <- as.matrix(GetAssayData(tumor_seurat, slot = "counts"))
tumor_clusters <- Idents(tumor_seurat)

# Extract count data and cluster information for healthy samples
healthy_counts <- as.matrix(GetAssayData(healthy_seurat, slot = "counts"))
healthy_clusters <- Idents(healthy_seurat)

# Function to calculate average expression per cluster
calculate_avg_expression <- function(counts, clusters) {
    avg_expression <- counts %>%
      as.data.frame() %>%
      rownames_to_column("gene") %>%
      pivot_longer(-gene, names_to = "cell", values_to = "expression") %>%
      left_join(data.frame(cell = colnames(counts), cluster = clusters), by = "cell") %>%
      group_by(gene, cluster) %>%
      summarize(avg_expression = mean(expression, na.rm = TRUE)) %>%
      pivot_wider(names_from = cluster, values_from = avg_expression)
    return(avg_expression)
}

# Calculate average expression for tumor samples
tumor_avg_expression <- calculate_avg_expression(tumor_counts, tumor_clusters)

# Calculate average expression for healthy samples
healthy_avg_expression <- calculate_avg_expression(healthy_counts, healthy_clusters)

# Export average expression to CSV
write.csv(tumor_avg_expression, file = "tumor_avg_expression.csv", row.names = TRUE)
write.csv(healthy_avg_expression, file = "healthy_avg_expression.csv", row.names = TRUE)
```


