---
title: "scRNA_annotation"
author: "Kristi Fourie"
date: "2024-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## scRNA Annotation of Tregs using SingleR

This document presents an analysis of scRNA-seq data focusing on regulatory T-cell identification and characterization using R packages such as Seurat, SingleCellExperiment, and SingleR.

## Load Libraries and Data
```{r load-libraries}
library(celldex)
library(SingleR)
library(scRNAseq)
library(Seurat)
library(SingleCellExperiment)
```

## Loadig preprocessed data 
```{r load-data}
ref <- MonacoImmuneData()
seuset_data <- readRDS("tumor_processed.rds") #change to healthy_processed.rds when running on healthy data
```

## Data Conversion: Seurat to SingleCellExperiment
```{r data-conversion}
sce_data <- as.SingleCellExperiment(seuset_data)
```

## Annotation using SingleR
```{r annotation}
results <- SingleR(test = sce_data, ref = ref, labels = ref$label.fine)
```

## Update Seurat Object with Annotations
```{r updated-seurat}
seuset_data <- AddMetaData(seuset_data, results$pruned.labels, col.name = "cell_type")
```

## Identify and Subset T Cells
```{r Tcell-subset}
t_cell_labels <- c("Central memory CD8 T cells", 
                   "Effector memory CD8 T cells",
                   "Follicular helper T cells",
                   "MAIT cells",
                   "Naive CD4 T cells",
                   "Naive CD8 T cells",
                   "Non-Vd2 gd T cells",
                   "T regulatory cells",
                   "Terminal effector CD4 T cells",
                   "Terminal effector CD8 T cells",
                   "Th1 cells",
                   "Th1/Th17 cells",
                   "Th17 cells",
                   "Th2 cells",
                   "Vd2 gd T cells")
pattern <- paste(t_cell_labels, collapse = "|")
t_cell_indices <- grep(pattern, seuset_data$cell_type)
t_cells_dataset <- seuset_data[, t_cell_indices]
```

## Dimensionality Reductions
```{r dimensionality-reduction}
# Perform PCA and UMAP if not already present
if (!"pca" %in% Reductions(t_cells_dataset)) {
  t_cells_dataset <- RunPCA(t_cells_dataset, verbose = FALSE)
}
if (!"umap" %in% Reductions(t_cells_dataset)) {
  t_cells_dataset <- RunUMAP(t_cells_dataset, reduction = "pca", dims = 1:20)
}
```

## Visualisation
This is done for validation of the Treg labels by visualising known Tregs markers
```{r visualisation}
DimPlot(t_cells_dataset, reduction = "umap", group.by = "cell_type", label = TRUE)
FeaturePlot(t_cells_dataset, features = c('FOXP3', 'CTLA4'))

# Color palette for cell types
all_cell_types <- unique(t_cells_dataset$cell_type)
cell_colors <- setNames(rep("grey", length(all_cell_types)), all_cell_types)
cell_colors["T regulatory cells"] <- "red" # Assign red color to Tregs

# UMAP plot with Tregs highlighted
DimPlot(t_cells_dataset, reduction = "umap", group.by = "cell_type", label = TRUE, cols = cell_colors)
```

## Quantification and Saving
Saving only the Tregs as RDS and the raw count matrix
```{r saving}
number_of_tregs <- sum(t_cells_dataset$cell_type == "T regulatory cells")

# Extract raw count matrix of Tregs
treg_cells <- subset(t_cells_dataset, cells = WhichCells(t_cells_dataset, expression = cell_type == "T regulatory cells"))
treg_counts <- GetAssayData(treg_cells, slot = "counts")

# Save raw count matrix as CSV
write.csv(as.matrix(treg_counts), file = "tumor_raw_counts.csv") #change to healthy_raw_counts.csv when running on healthy data

# Save the Seurat object
saveRDS(t_cells_dataset, file = "tumor_tregs.rds") #change to healthy_tregs.rds when running on healthy data
```


